---
title: 'HOLC Grade Inequality'
author: "Ben Versteeg"
output: 
  html_document:       
    print_df: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
    self_contained: false
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
warning: False
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE, echo = TRUE)
getwd() 
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Data

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

## Investigating the legacy of redlining in current environmental (in)justice

### Importing data and libraries

```{r}
# Read in necessary packages
library(tidyverse)
library(sf)
library(tmap)
```

Read in EJScreen data and filter to Los Angeles County

```{r include=TRUE, warning=FALSE}
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb") 

california <- ejscreen %>%
  filter(ST_ABBREV == "CA") 

# filter to a county you are interested in
la_county <- california %>%
  filter(CNTY_NAME %in% c("Los Angeles County"))
```

### Data Analysis

#### Wastewater discharge

Let's create a map of wastewater discharge by census block groups.
Indicate which census block groups are above the 95th percentile of
national values for wastewater discharge by adding a centroid.

```{r}
# Filtering for centroids above the 95th percentile
la_centroid <-  st_centroid(la_county) %>% 
  filter(P_PWDIS > 95)


# Map 1: Base map with filled census block groups representing wastewater discharge
map1 <- tm_shape(la_county) + # Create a base map using LA County geometry
  tm_borders() + # Add borders to the map
  tm_fill("PWDIS", # Fill census block groups based on "PWDIS" variable
          style = "jenks", # Jenks classification
          n = 7, palette = "viridis",  # 7 breaks, viridis palette
          title = "Wastewater Discharge") + # Add a legend title for the fill
  tm_legend(outside = TRUE) # Place the legend outside the map

# Map 2: Build upon Map 1, adding red dots at the centroids of census block groups
map2 <- map1 +
  tm_shape(la_centroid) + # Add a new layer with centroids of census block groups
    tm_dots(size = 0.10, aes(color = "red"), # Display dots at centroids, red in color
            fill_alpha = 1) + # Make the dots fully opaque
    tm_add_legend(type = "fill", # Add a legend indicating red dots
                  title = "Census Block Groups", # Legend title
                  labels = c("Above 95 percentile"), # Legend labels
                  fill = "red") + # Color for the legend
    tm_layout(main.title = "Wastewater Discharge By Census Block Group", # Set main title
              xlab.text = "Longitude", # X-axis label
              ylab.text = "latitude") + # Y-axis label
    tm_graticules() + # Add graticules (latitude and longitude lines)
    tm_scale_bar() + # Add a scale bar
    tm_compass(position = c("left", "bottom"), # Add a compass in the bottom-left corner
               size = 1) # Set the compass size

# Print Map 2
map2


```

This map is of LA County and it shows the percentiles of wastewater
discharge. The least amount of discharge is represented by green and the
most discharge is represented by dark purple. The red areas are cencus
block groups over the 95th percentile.

#### Low Income Percentage

Next, we'll find the percent of census block groups that have:\
- less than 5% of the population is considered low income

```{r}
# filtering for low income percent that is less than 5 percent
la_centroid1 <- st_centroid(la_county) %>% 
  filter(LOWINCPCT < .05)

# divide the number of rows in la_centroid1 by rows in la_county
nrow(la_centroid1) / nrow(la_county)

# 6%
```

In this analysis, we observe a proportion of approximately 6%. One
inference drawn from this figure suggests a notable presence of census
blocks with a greater share of residents classified as having a low
income. This implies that there may be a higher concentration of
individuals with lower income levels residing in Los Angeles County.
Further exploration of the data is warranted to determine if this
low-income population is more prevalent in areas assigned lower HOLC
ranks.

#### Particulate Matter and Superfund proximity

Having examined the percentage of low-income residents across various
census block groups, we will now investigate the association between
census block groups and their proximity to Particulate Matter and
Superfund sites.

We will find the percent of census block groups that are:\
- above the 80th percentile for Particulate Matter 2.5 AND\
- above the 80th percentile for Superfund proximity

```{r}
# filtering above the 80 percentile
la_centroid2 <- st_centroid(la_county) %>% 
  filter(P_PM25 > 80 & P_PNPL > 80) 

nrow(la_centroid2) / nrow(la_county)

# 17%
```

Based on this analysis, it can be inferred that approximately 17% of
census block groups reside in regions that fall within the 80th
percentile concerning elevated PM 2.5 levels and proximity to Superfund
sites.

#### Import redlining information for Los Angeles.

Read in the redlining data for Los Angeles county from a URL link.

```{r}
# Reading in redlining data
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid()

```

#### HOLC Grade LA County Map

Let's make a map of historical redlining boundaries, colored by HOLC
grade.

```{r}
# Map layer 1: Historical Redlining Boundaries
tm_shape(LA_redlining) + # Define the base layer using the LA_redlining dataset
  tm_graticules() + # Add graticules (latitude and longitude lines)
  tm_polygons(fill = "grade", col = "black") + # Add polygons filled based on the "grade" variable, outlined in black
  tm_shape(la_county) + # Add a new layer with LA County boundaries
  tm_borders(col = "darkgrey", col_alpha = 0.2) + # Add county borders in dark grey with 20% transparency
  tm_title("Historical Redlining Boundaries") + # Set the main title for the map
  tm_scalebar(position = c("left", "bottom")) + # Add a scale bar in the bottom-left corner
  tm_compass(position = c("right", "bottom")) # Add a compass in the bottom-right corner

```

This figure shows us where each HOLC grade is throughout LA County.

#### HOLC Grade Census Block Groups

Let's now find the number of census block groups that fall within areas
with HOLC grades

```{r eval=FALSE}
st_crs(LA_redlining) <- st_crs(la_county)
  
st_crs(la_county) == st_crs(LA_redlining)

la_join <- st_join(la_county, LA_redlining)

nrow(la_join)
```

This tells us that there are a total of 6591 census block groups that
fall within HOLC grade areas.

#### Summarizing Redlining data

Let's summarize the current conditions based on EJScreen data within
historical redlining categories using the mean of the following
variables:\
 - % low income.\
 - percentile for particulate Matter 2.5.\
 - percentile for low life expectancy.\
 - percentile for air toxics cancer risk (20 points)

```{r}
mean(ejscreen$LOWINCPCT, na.rm = TRUE) * 100
mean(ejscreen$P_PM25, na.rm = TRUE)
mean(ejscreen$LIFEEXPPCT, na.rm = TRUE) * 100
mean(ejscreen$P_CANCER, na.rm = TRUE)
```

The mean percent of low income is 30%, the mean percentile for
particulate matter 2.5 is 50%, the mean percentile for low life
expectancy is 19%, and the mean percentile for air toxic cancer risk is
24%. Based on this analysis, individuals residing in areas assigned an A
HOLC grade experience a higher standard of living compared to those in
areas with a D HOLC grade. The A HOLC grade is associated with a larger
income pool, implying greater access to quality healthcare and a cleaner
living environment, in contrast to areas with lower income grades.

## Investigating the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC
grads, let's find the percent of observations within each redlining
categories and plot results.

```{r}
# Read in the birds data
gbif_birds <- st_read("data/gbif-birds-LA") 
```

```{r}
# Set the CRS of gbif_birds to match LA_redlining
st_crs(gbif_birds) <- st_crs(LA_redlining)

# Spatial join based on intersection between LA_redlining and gbif_birds
joined_birds <- st_join(LA_redlining, gbif_birds, join = st_intersects)

# Filter for only bird records from the year 2022
birds <- joined_birds[joined_birds$year == 2022, ]

# Count the number of bird records
nrow(birds)

# Calculate the percentage of bird records for each HOLC grade
birds_sum <- birds %>% 
  group_by(grade) %>% # Group data by HOLC grade
  summarise(birds_p = n()/nrow(birds)*100) %>% # Calculate the percentage of bird records for each grade
  st_drop_geometry() # Drop the spatial geometry for summary purposes

birds_sum
```

```{r}
# Create a bar plot using birds_sum data
ggplot(data = birds_sum, aes(x = grade, y = birds_p)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    x = "Historical Redlining Categories", # X-axis label
    y = "Percentage of Bird Observations", # Y-axis label
    title = "Distribution of Bird Observations by Redlining Category (Year 2022)"
  ) + # title
  theme_minimal()  # Set the theme to minimal


```

Based on the results, it seems that there was a higher percentage of
bird observations in areas with lower HOLC grades, specifically, the C
category. It is surprising that there would be more birds in areas with
lower HOLC grades but there could be a few explanations on why this is
the case. Areas with lower HOLC grades may have retained more natural
habitats, such as parks, open spaces, and undeveloped land. There could
also be more food sources and ecological diversity. Another reason is
that there could be some sampling bias where there could be more
birdwatchers and researchers in lower HOLC grade areas which results in
more observations in those areas, even if the bird populations are not
actually higher.
